Pigeons.jl: Distributed sampling from intractable distributions

### 2023-08-18 src/SPI.tex
Minor changes before submission

\medskip 
To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
are exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in the reduction tree, irrespective of the 
number of machines available to perform the reduction. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
illustrate the reduction procedure for eight and two machines, respectively.

@@

\subsection{Splittable random streams}
\label{sec:splittable_randoms}
Another building block towards achieving SPI is a \emph{splittable random stream}
\cite{lecuyer1988splittable,burton1992splittable}. 
Julia uses \emph{task-local} random number generators, a notion that is 
related to (but does not necessarily imply) strong parallelism invariance. 
A \emph{task} is a unit of work on a machine.
A task-local RNG would then mean that a separate RNG is 
used for each unit of work, hopefully implying strong parallelism invariance
if the number of tasks is assumed to be constant. 
Unfortunately, this is not the case when a separate task is created for each thread 
of execution in Julia.
We note that the \texttt{@threads} macro in Julia creates exactly \texttt{nthreads()}
tasks and thus \texttt{nthreads()} pseudorandom number generators.
This can break strong parallelism invariance as the output may depend on the number of 
threads.

>>>>

\medskip 
To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
is exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in the reduction tree, irrespective of the 
number of machines available to perform the reduction. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
illustrate the reduction procedure for eight and two machines, respectively.

@@

\subsection{Splittable random streams}
\label{sec:splittable_randoms}
Another building block towards achieving SPI is a \emph{splittable random stream}
\cite{lecuyer1988splittable,burton1992splittable}. 
Julia uses \emph{task-local} random number generators, a notion that is 
related to (but does not necessarily imply) strong parallelism invariance. 
A \emph{task} is a unit of work on a machine.
A task-local RNG would then mean that a separate RNG is 
used for each unit of work, hopefully implying strong parallelism invariance
if the number of tasks is assumed to be constant. 
Unfortunately, this is not the case when a separate task is created for each thread 
of execution in Julia.
We note that the \texttt{@threads} macro in Julia creates \texttt{nthreads()}
tasks and thus \texttt{nthreads()} pseudorandom number generators.
This can break strong parallelism invariance as the output may depend on the number of 
threads.

### 2023-08-18 src/examples.tex
Minor changes before submission

\subsection{Targets}
To use Pigeons.jl, we must specify a target distribution, given by $\gamma$ 
in \cref{eq:normalizing_constant}.
Numerous possible types of target distributions are supported, including 
custom probability densities (specified up to a normalizing constant) written in Julia.
We also allow to interface with models written in common probabilistic programming 
languages, including:
\begin{itemize}
    \item Turing.jl \cite{ge2018turing} models (\texttt{TuringLogPotential})
    \item Stan \cite{carpenter2017stan} models (\texttt{StanLogPotential})
    \item Comrade.jl models for black hole imaging (\texttt{ComradeLogPotential})
    \item Non-Julian models with foreign-language Markov chain Monte Carlo (MCMC) code 
    (e.g. Blang \cite{bouchard2022blang} code for phylogenetic inference over combinatorial spaces) 
\end{itemize}
Additional targets are currently being accommodated and will be 
introduced to Pigeons.jl in the near future.

@@

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.jl.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

@@

    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
    log of the normalization constant. See \cref{sec:PT_diagnostics} for more information.

>>>>

\subsection{Targets}
To use Pigeons.jl, we must specify a target distribution, given by $\gamma$ 
in \cref{eq:normalizing_constant}.
Numerous possible types of target distributions are supported, including 
custom probability densities (specified up to a normalizing constant) written in Julia.
We also allow to interface with models written in common probabilistic programming 
languages, including:
\begin{itemize}
    \item Turing.jl \cite{ge2018turing} models (\texttt{TuringLogPotential})
    \item Stan \cite{carpenter2017stan} models (\texttt{StanLogPotential})
    \item Comrade.jl\footnote{\url{https://github.com/ptiede/Comrade.jl}} 
      models for black hole imaging (\texttt{ComradeLogPotential})
    \item Non-Julian models with foreign-language Markov chain Monte Carlo (MCMC) code 
    (e.g. Blang \cite{bouchard2022blang} code for phylogenetic inference over combinatorial spaces) 
\end{itemize}
Additional targets are currently being accommodated and will be 
introduced to Pigeons.jl in the near future.

@@

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

@@

    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
    logarithm of the ratio of normalization constants between the target and the reference. 
    In many cases, $Z_0 = 1$.
